---
title: "Covid"
author: "Budha Gautam"
date: "2025-07-23"
output: html_document
---

```{r setup, include=FALSE}
library(stringr)
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

```{r get_jhu_data}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c(
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_deaths_US.csv"
)

urls <- str_c(url_in, file_names)
```

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

```

```{r tidy_global_data , error=TRUE}
global_cases <- global_cases %>%
  pivot_longer(
    cols = -c(`Province/State`, `Country/Region`, Lat, Long),
    names_to = "date",
    values_to = "cases")%>%
  select(-c(Lat, Long))


global_deaths <- global_deaths %>%
  pivot_longer(
    cols = -c(`Province/State`, `Country/Region`, Lat, Long),
    names_to = "date",
    values_to = "deaths")%>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
    rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>%
    mutate(date = mdy(date))

global <- global %>% filter(cases > 0)



global %>% filter(cases > 103000000)

us_cases

us_cases <- us_cases %>% 
  pivot_longer(cols = -(UID:Combined_Key), 
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2: cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
  

us_deaths <- us_deaths %>% 
  pivot_longer(cols = -(UID:Population), 
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2: deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
  
US <- us_cases %>%
  full_join(us_deaths)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region), 
        sep = ", ", 
        na.rm = TRUE, 
        remove = FALSE)

uid_lookup_url <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")

uid <- uid_lookup_url %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% 
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS))%>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)


```

```{r tidy_us_data_vis , error=TRUE}

US_by_State <- US %>%
  group_by(Province_State, Country_Region, date)%>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population))%>%
  mutate(deaths_per_mill = deaths *1000000 / Population)%>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()


US_totals <- US_by_State %>%
  group_by( Country_Region, date)%>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population))%>%
  mutate(deaths_per_mill = deaths *1000000 / Population)%>%
  select( Country_Region, date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()


tail(US_totals)

US_totals %>%
  filter(cases> 0)%>%
  ggplot(aes(x= date, y = cases)) + 
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10() + 
  theme(legend.position = "buttom", 
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US" , y = NULL)
  
state <- "New York"
US_by_State %>%
  filter(Province_State == state, cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) +
  labs(title = paste("COVID-19 in", state), y = NULL)

max(US_totals$date)
max(US_totals$deaths)

US_by_State <- US_by_State  %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals  %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals)
tail(US_totals %>% select(new_cases, new_deaths, everything()))


```{r top5_new_deaths}
latest_date <- max(US_by_State$date, na.rm = TRUE)

US_by_State %>%
  filter(date == latest_date) %>%
  arrange(desc(new_deaths)) %>%
  slice_head(n = 5) %>%
  ggplot(aes(x = reorder(Province_State, -new_deaths), y = new_deaths)) +
  geom_col(fill = "red") +
  labs(title = paste("Top 5 States by New Deaths on", latest_date),
       x = "State", y = "New Deaths") +
  theme_minimal()


#new total
US_totals %>%
  filter(cases> 0)%>%
  ggplot(aes(x= date, y = new_cases)) + 
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = deaths, color = "new_deaths"))+
  geom_point(aes(y = deaths, color = "new_deaths"))+
  scale_y_log10() + 
  theme(legend.position = "buttom", 
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US" , y = NULL)
  

state <- "New York"
US_by_State %>%
  filter(Province_State == state, cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = deaths, color = "new_deaths")) +
  geom_point(aes(y = deaths, color = "new_deaths")) +
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) +
  labs(title = paste("COVID-19 in", state), y = NULL)

    
US_state_totals <- US_by_State%>%
  group_by(Province_State)%>%
  summarize(deaths = max(deaths), cases = max(cases), Population = max(Population), cases_per_thou = 1000*cases / Population, deaths_per_thou = 1000* deaths / Population)%>%
  filter(cases>0, Population> 0)


US_state_totals%>%
  slice_min(deaths_per_thou, n = 10)

US_state_totals%>%
  slice_min(deaths_per_thou, n = 10)%>%
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals%>%
  slice_max(deaths_per_thou, n = 10)%>%
  select(deaths_per_thou, cases_per_thou, everything())


```{r case_fatality_rate}
US_state_totals %>%
  mutate(CFR = deaths / cases * 100) %>%
  filter(cases > 10000) %>%  # Filter out small case counts
  arrange(desc(CFR)) %>%
  slice_max(CFR, n = 10) %>%
  ggplot(aes(x = reorder(Province_State, -CFR), y = CFR)) +
  geom_col(fill = "purple") +
  labs(title = "Top 10 States by Case Fatality Rate",
       x = "State", y = "CFR (%)") +
  theme_minimal()


```

```{r tidy_us_pred_data , error=TRUE}

mod<- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)

US_state_totals %>% slice_min(cases_per_thou)

US_state_totals %>% slice_max(cases_per_thou)

#x_grid <- seq(1, 151)

#new_df <- tibble(cases_per_thou = x_grid)

US_state_totals%>% mutate(pred = predict(mod))

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred

US_tot_w_pred %>% ggplot() +
  geom_point(aes(x= cases_per_thou, y = deaths_per_thou), color = "blue" )+
  geom_point(aes(x= cases_per_thou, y = pred), color = "red" )

```




### Potential Biases in the Data and Analysis

COVID-19 data may not fully reflect reality due to limited testing and delays in reporting. For example, cases and deaths are often undercounted—especially in regions with fewer resources or during periods of overwhelmed healthcare systems.

Daily numbers can also fluctuate because of delays in reporting over weekends or holidays, leading to sudden dips or spikes in the data.

Our model assumes that all U.S. states are comparable, but that’s rarely the case. Factors like the age of the population, access to healthcare, urban versus rural settings, and economic differences can all impact the outcomes in significant ways.

Finally, the data covers a long time period during which many important things changed—such as mask mandates, lockdowns, vaccine rollouts, and shifts in public behavior. These changes can greatly influence the trends but are not captured in our model.
