---
title: "NYPD Shooting"
author: "Budha Gautam"
date: "2025-07-13"
output: 
  pdf_document:
    toc: true
    number_sections: true
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(DT)
```
## NYPD Shooting Overview

This report investigates New York City shooting incidents from 2006 through 2024. The dataset captures each shooting’s date, time, borough, location classification, and demographic details for both victims and suspects.  

My goal is to develop a logistic regression model that estimates the probability that a given shooting resulted in a homicide, using features such as the borough in which the shooting occurred and the month it occurred. 

Key tools in this analysis:
- **tidyverse** for data cleaning, transformation, and plotting  
- **caret** for partitioning the data and computing the confusion matrix  

```{r data-loading}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_cases <- read_csv(url(url_in))

nypd_cases <- nypd_cases %>%
  mutate(
    VIC_AGE_GROUP = case_when(
      VIC_AGE_GROUP == "1022" ~ NA_character_,
      TRUE ~ VIC_AGE_GROUP
    )
  )

print(head(nypd_cases, 10))

```

```{r date-processing}
nypd_cases <- nypd_cases %>%
  mutate(
    occur_date = as.Date(OCCUR_DATE, "%m/%d/%Y"),
    year = as.integer(format(occur_date, "%Y")),
    month = as.Date(paste0(format(occur_date, "%Y-%m"), "-01"))
  ) %>%
  select(-OCCUR_DATE)

head(nypd_cases[c("occur_date", "year", "month")])
```

```{r monthly-trends}
monthly_counts <- nypd_cases %>% count(month)
ggplot(monthly_counts, aes(month, n)) +
  geom_line() +
  labs(title = "Monthly Shooting Incidents", x = "Month", y = "Count") +
  theme_minimal()
```

```{r borough-analysis}
boro_counts <- nypd_cases %>% 
  filter(!is.na(BORO)) %>% 
  count(BORO)

ggplot(boro_counts, aes(BORO, n, fill = BORO)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Shooting Incidents by Borough", x = "Borough", y = "Count") +
  theme_minimal()
```

```{r modeling}
model_data <- nypd_cases %>%
  filter(
    !is.na(STATISTICAL_MURDER_FLAG),
    !is.na(BORO),
    !is.na(VIC_AGE_GROUP),
    !is.na(VIC_RACE)
  ) %>%
  transmute(
    murder = STATISTICAL_MURDER_FLAG,
    BORO, 
    vic_age = VIC_AGE_GROUP, 
    vic_race = VIC_RACE
  ) %>%
  mutate(across(everything(), as.factor))

set.seed(123)
train_idx <- createDataPartition(model_data$murder, p = 0.7, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

test_data <- test_data %>%
  mutate(
    vic_age = factor(vic_age, levels = levels(train_data$vic_age)),
    vic_race = factor(vic_race, levels = levels(train_data$vic_race)),
    BORO = factor(BORO, levels = levels(train_data$BORO))
  )

model <- glm(murder ~ BORO + vic_age + vic_race,
             data = train_data, 
             family = binomial)

summary(model)
```

```{r model-evaluation}
pred_probs <- predict(model, test_data, type = "response")
pred_class <- factor(pred_probs > 0.5, levels = c(FALSE, TRUE))

confusionMatrix(pred_class, test_data$murder)
```
# Conclusion

The logistic regression model achieved an overall accuracy of **79.9%**.

- **Sensitivity (non-homicide detection)**: 100%  
- **Specificity (homicide detection)**:   0%  

This happens because only about 20% of incidents are flagged as homicides, so predicting the majority class yields ~80% accuracy.  

**Key takeaways & biases**  
- **Class imbalance**: Homicides (~20%) vs. non-homicides (~80%) skews the model.  
- **Limited features**: Borough + victim demographics alone lack predictive power.  
- **Reporting bias**: NYPD logging practices may under- or over-represent certain shootings.  
